---
title: "Continuous culture model for phytoplankton"
author: "Andrew Irwin"
date: "2020 July 23"
output: 
  html_document:
    toc: true
    toc-float: true
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(deSolve)
library(ggplot2)
library(tidyr)
library(dplyr)
library(forcats)
library(plotly)
```

## Overview

In this document I explore a variable internal stores model of phytoplankton culture growth adapted from [Grover (1991)](https://www.journals.uchicago.edu/doi/abs/10.1086/285254).

There are three pools: resources in the growth medium, such as nitrate, $R$ (mol L$^{-1}$), resources stored in the cell and available for growth, $Q$ (mol cell$^{-1}$), and the cell density, $X$ (cells L$^{-1}$). In the examples below I don't use these units. Instead, all quantities are chosen to be close to 1. Plausible values for _Thalassiosira pseudonana_ and the development of non-dimensionalized equations are explored after some examples. Finally I discuss the connection between these models and estimating growth rates in culture.

## Equations

The equations for the three pools are

$$\frac{dR}{dt} = d(R_s - R) - \rho X$$
$$\frac{dQ}{dt} = \rho - \mu Q $$
$$\frac{dX}{dt} = X(\mu-d).$$

Which can be read as the following: (1) Resources in the media are supplied at a rate $d$ and concentration $R_s$, are consumed at a rate proportional to the uptake $\rho$ of each cell times the density of cells $X$, and unconsumed resources are washed out at a rate $d$. (2) Internal stores of the nutrient for each cell are increased at a rate $\rho$ and depleted at the growth rate $\mu$ times the current amount of storage per cell $Q$. (3) The cell density increases at the growth rate $\mu$ and decreases at the rate of dilution $d$. Initial values at time $t=0$ will be written as $R_0, Q_0,$ and $X_0$.

The uptake and growth functions are defined as 

$$\rho(R) = \frac{V_{max}R}{K_m+R}$$
$$\mu(Q) = \mu_{max}\left(1-\frac{Q_{min}}{Q}\right).$$
Both functions are saturating functions: $V_{max}$ and $K_m$ define the shape of the Michaelis-Menten uptake curve and $\mu_{max}$ and $Q_{min}$ define the shape of the growth response which is $0$ when internal stores are depleted to the minimum value $Q_{min}$.

Total mass should change smoothly from the initial value to a value determined by the supply concentration $R_s$ and can be computed as 

$$M = R + XQ$$
If the initial mass in the reaction vessel is very close to the equilibrium value then very small and noisy deviations will be visible in the trajectory of $M$.

A commonly used parameter from the simpler Monod model (which has no cell quota $Q$ and writes the growth function the way we wrote the updake function) is the minimum resource concentration needed for the cells to grow. This value $R^*$ can be found from the equilibrium solution with $X>0$ by 
setting the derivatives equal to 0:

$$R^* = \frac{K_m d}{\mu_{max}-d}.$$
Note that as the dilution rate increases to approach the maximum growth rate, the concentration of resource required to support these cells increases without bound.

For the Droop model, the algebra and corresponding expression is somewhat more complicated, since we need to find both $Q^*$ and $R^*$ from $\frac{dX}{dt} = 0$ and $\frac{dQ}{dt}=0$:

$$Q^* = \frac{\mu_{max}Q_{min}}{\mu_{max}-d}$$

$$R^* = \frac{\mu_{max}K_m(Q^*-Q_{min})}{V_{max}-\mu_{max}(Q^*-Q_{min})} =  \frac{K_m}{\frac{V_{max}}{\mu_{max}(Q^*-Q_{min})}-1}$$
We notice several things from these expressions. In a batch culture, with $d=0$, $Q^* = Q_{min}$ and $R^* = -K_m$. (We interpret a negative $R^*$ as meaning that $R$ can be drawn down to 0.) In a chemostat, as we increase $d$ towards $\mu_{max}$, $Q^*$ grows ever larger and $R^*$ is eventually $0$. In a turbidostat, the dilution rate is adjusted to match the growth rate, and we obtain $R^*$ from $\frac{dR}{dt}=0$ yielding the rather unwieldy expression

$$R^* = \frac{1}{2}\left( R_s-K_m-\frac{X^*V_{max}}{d}+\sqrt{(R_s-K_m-X^*V_{max}/d)^2+4R_sK_m}\right)$$

where $d = \mu = \mu_{max}\left(1-\frac{Q_{min}}{Q^*}\right)$ and $X^*$ is set by the experimenter. (We selected the positive root from the quadratic formula.)






```{r}
# system of equations

# dR/dt = d(R0 - R) - rho(Q)X
# dQ/dt = rho(Q) - mu*Q
# dX/dt = X (mu(Q) - d)
# mu(Q) = mumax ( 1 - Qmin/Q)
# rho(R) = Vmax R / (Km + R)
# conservation law: d/dt(R + XQ) = 0 = dR/dt + XdQ/dt + QdX/dt 

culture.chemostat <- function(t, state, parameters) {
    with(as.list(c(state, parameters)), {
        rho = Vmax*R/(Km+R)
        mu = mumax*(1-Qmin/Q)
        dR = d*(Rs - R) - rho*X
        dQ = rho - mu*Q
        dX = X*(mu-d)
        list(c(dR, dQ, dX), c(rho=rho, mu=mu, mass = R + Q*X))  
    })
}

# in a turbidostat, we dilute at 2*mumax whenever X > Xstar; maybe use sigmoid function to avoid discontinuities
culture.turbidostat <- function(t, state, parameters) {
    with(as.list(c(state, parameters)), {
        rho = Vmax*R/(Km+R)
        mu = mumax*(1-Qmin/Q)
        # d = 2*mumax/(1+exp(-4*(X-Xstar)/X))
        d = ifelse(X>Xstar, 2*mumax, 0)  # discontinuity; hard to integrate
        d = case_when( X < 0.95*Xstar ~ 0,
                       X < 1.05*Xstar ~ 2*mumax*(X-0.95*Xstar)/(0.10*Xstar), 
                       TRUE ~ 2*mumax)
        dR = d*(Rs - R) - rho*X
        dQ = rho - mu*Q
        dX = X*(mu-d)
        list(c(dR, dQ, dX), c(rho=rho, mu=mu, mass = R + Q*X))  
    })
}

# sample integration
# params <- c(mumax = 1.0, Vmax = 2.0, Km = 1.0, Qmin = 0.1, Rs = 2.0, d = 0.5)
# state <- c(R = 1.0, Q = 1.0, X = 1.0)

# culture.chemostat(0, state, params)
# times <- seq(0, 25, length = 100)
# out <- ode(y = state, times = times, func = culture.chemostat, parms = params) %>% as.data.frame()

plotTS <- function(out) {
  out %>% as_tibble() %>%
    mutate(log10X = log10(X)) %>%
    select(-X) %>%
    pivot_longer(R:log10X) %>%
    mutate(name = fct_relevel(name, "R", "Q", "log10X", "rho", "mu","mass")) %>%
    ggplot(aes(x=time, y = value )) + 
    geom_line(size=2) + 
    expand_limits(y=0) +
    facet_wrap( vars(name), scale="free_y") + 
    theme_bw() + 
    theme(axis.text = element_text(size=14),
          axis.title = element_text(size=14), 
          strip.text = element_text(size=14)) 
}

plotTS2 <- function(out) {
  o2 <- as.data.frame(out)
  o2$log10X = log10(o2$X)
  vars <- c("R", "rho", "Q", "mu", "log10X", "mu") #  setdiff(names(o2), "time")
  plots <- lapply(vars, function(var) {
    plot_ly(o2, x = ~ time, y = as.formula(paste0("~", var))) %>%
      add_lines(name=var)
  })
  subplot(plots, nrows=length(plots)/2, shareX=TRUE, titleX=FALSE) %>% 
    layout(legend = list(orientation = 'h')) # %>%
    # layout(xaxis = list(title="Time"))
}
# plotTS(out)
```

## Examples

### Batch culture

Fresh medium inoculated with a culture and allowed to grow as a closed system -- with no material added or removed -- is known as a batch culture. The general trajectory is that resources are consumed and cells grow until resources become limiting at which point growth slows and eventually ceases. This is the most straightforward experimental setup and also roughly corresponds to the trajectory of a seasonal phytoplankton bloom.

For a batch culture, we set the dilution rate $d=0$ and choose all the initial conditions. Here I have set $\mu_{max} = 1.0$ with a maximum uptake rate of twice this $V_{max} = 2.0$, the half-saturation constant for resource uptake is quite small $K_m=1.0$ compared to the supply concentration $R_0=2.0$. The minimum quota $Q_{min}$ is small compared to the initial value of cell quota $Q=1.0$, and the initial resource concentration in the media is $R=1.0$, so the cells are initially resource replete. The initial cell densitity is $X=1.0$ so the total mass is $M = 2$ and the equilibrium mass is $R_0=2$.

A culture acclimated to the initial conditions (so that $\rho = \mu Q$ at time 0) will grow exponentially at time 0 with a gradually decreasing growth rate until resources are depleted. The equilibrium solution will have extracellular resources drawn down to 0, intracellular storage drawn down to $Q_{min}$ and a zero growth rate. A real culture will exhibit some cell death and decrease in cell density, but that is not included in our model.

```{r}
# sample integration
params <- c(mumax = 1.0, Vmax = 2.0, Km = 1.0, Qmin = 0.1, Rs = 2.0, d = 0)
state <- c(R = 1.0, Q = 1.0, X = 1.0)

# culture.chemostat(0, state, params)
times <- seq(0, 10, length = 100)
out <- ode(y = state, times = times, func = culture.chemostat, parms = params) %>% as.data.frame()
plotTS(out)
```

### Chemostat culture

The opposite extreme to a batch culture is the chemostat which undergoes dilution with fresh media at a constant (non-zero) rate $d$ and after some time achieves an equilibrium uptake rate, cell storage, growth rate, and cell density. Cells will acclimate to grow at the dilution rate so that $\mu=d$ and we say that the dilution rate sets the growth rate. These experiments require pumps and a steady supply of fresh media to keep the cells growing. They are a kind of gold-standard for ensuring that equilibrium conditions are established. If the dilution rate is too high (close to $\mu_{max}$) the culture will be susceptible to washout or catastrophic failure. If the dilution rate is very slow, the growth rate will be slow, but perhaps not due to the limitation of the resource described in the equations; other factors such as DIC (pH) or light might be limiting growth.

Here is an example with $d = 0.5 = \mu_{max}/2$ and otherwise all conditions the same as the batch culture experiment. Note in particular that at equilibrium growth rate is equal to the
imposed dilution rate.

```{r}
# sample integration
params <- c(mumax = 1.0, Vmax = 2.0, Km = 1.0, Qmin = 0.1, Rs = 2.0, d = 0.5)
state <- c(R = 1.0, Q = 1.0, X = 1.0)

# culture.chemostat(0, state, params)
times <- seq(0, 20, length = 100)
out <- ode(y = state, times = times, func = culture.chemostat, parms = params) %>% as.data.frame()
plotTS(out)
```

### Turbidostat culture

Changing the resource supply rate in a chemostat does not affect growth rate, since growth rate is determined by the dilution rate. An alternative continuous culture setup is the turbidostat which maintains constant cell numbers (assayed by turbidity in an experimental setup). The dilution rate is adjusted to match growth rate with the goal of maintaining a constant cell density. We define $d$ in our equations to be 0 if $X<X^*$ and a minimum of $\mu_{max}$ otherwise. (Numerically we do something a bit different since this defiition is not continuous and creates computational difficulties.)

This setup uses the same parameters as the chemostat system, but with $d$ set as described and a biomass density target of $X^* = 2.0$ (or $\log_{10}X^* = 0.30$).

```{r}
# sample integration
params <- c(mumax = 1.0, Vmax = 2.0, Km = 1.0, Qmin = 0.1, Rs = 2.0, Xstar = 2.0)
state <- c(R = 1.0, Q = 1.0, X = 1.0)

# culture.chemostat(0, state, params)
times <- seq(0, 5, length = 100)
out <- ode(y = state, times = times, func = culture.turbidostat, parms = params) %>% as.data.frame()
plotTS(out)
```

### Semi-continuous batch culture

A very common experimental setup in the lab combines features of these different arrangements. The goal is to have a simple setup that doesn't require a pump and media reservoir, but to maintain near-constant or at least near-ideal conditions for growth. The solution is to have a series of batch cultures which are diluted into fresh media before the cell density gets too high or the resources in the media are drawn down too low. The main goal is to have nearly constant growth rates.

We accomplish this with our equations by setting dilution rate to 0 and diluting periodically with a ratio of old culture and fresh media to achieve a fixed initial culture density $X^*$. In practice, instead of dilution every $T$ time units, it would be more conventional to dilute the day before $X$ was predicted to get above a specified threshold.

Here we use the same conditions as above and dilute the culture every $T=3$ time periods and I've increased the resource concentration in the supplied media to $R_s=30$ so that resources never get drawn down enough to be strongly limiting. Notice that $R$ and $\rho$ vary a lot, $Q$ much less, and $\mu$ very little.

```{r}
# sample integration
params <- c(mumax = 1.0, Vmax = 2.0, Km = 1.0, Qmin = 0.1, Rs = 30.0, d = 0, t_final = 15)
state <- c(R = 1.0, Q = 1.0, X = 1.0)

# generate model output
t0 = 0
T = 3
Xstar = 1.0
times <- seq(t0, t0+T, length = 100)
out <- ode(y = state, times = times, func = culture.chemostat, parms = params) %>% as.data.frame()
while(out$time[nrow(out)] < params["t_final"]) {
  t0 <- out$time[nrow(out)]
  times <- seq(t0, t0+T, length = 100)
  Df = Xstar/out$X[nrow(out)]
  state <- c(R = Df * out$R[nrow(out)] + (1-Df)*params["Rs"],
             Q = out$Q[nrow(out)],
             X = Df * out$X[nrow(out)] + 0) 
  names(state)[1] = "R"
  out <- bind_rows( out, ode(y = state, times = times, func = culture.chemostat, parms = params) %>% as.data.frame() )
}
        
# plot time series
plotTS(out)
```


## Interactive example

Here you can experiment with these equations, changing the settings and the obtaining the updated trajectories.


```{r}
commonParameters <- function() {
 flowLayout(
   sliderInput("t_final", "t_final:", min = 1, max = 100, value = 25.00, step = 1.0),
   sliderInput("mumax", "mumax:", min = 0, max = 2, value = 1.00, step = 0.05),
   sliderInput("Vmax", "Vmax:", min = 0, max = 2, value = 2.00, step = 0.05),
   sliderInput("Km", "Km:", min = 0, max = 2, value = 1.00, step = 0.05),
   sliderInput("Qmin", "Qmin:", min = 0, max = 2, value = 0.10, step = 0.05),
   sliderInput("Rs", "Rs:", min = 0, max = 50, value = 2.00, step = 0.05),
   sliderInput("R", "R0:", min = 0, max = 2, value = 1.00, step = 0.05),
   sliderInput("Q", "Q0:", min = 0, max = 2, value = 1.00, step = 0.05),
   sliderInput("X", "X0:", min = 0, max = 2, value = 1.00, step = 0.05)
 )
}
```
```{r}
w = 600; h = 400
fluidPage(
  sidebarLayout(
    sidebarPanel(
      verticalLayout(
        conditionalPanel(
          'input.select === "Chemostat"',
          verticalLayout(
             sliderInput("d", "d:", min = 0, max = 2, value = 0.50, step = 0.05)
          )        
        ),
        conditionalPanel(
          'input.select === "Turbidostat"',
          verticalLayout(
             sliderInput("log10Xstar", "log10 X*:", min = 0, max = 2, value = 0.50, step = 0.05)
          )        
        ),
        conditionalPanel(
          'input.select === "Semi-continuous"',
          verticalLayout(
             sliderInput("Period", "Period:", min = 5, max = 100, value = 20, step = 0.5),
             sliderInput("Xstar", "X*:", min = 0, max = 1, value = 0.50, step = 0.05)
          )        
        ),
        commonParameters()
      )
    ),
    mainPanel(
      tabsetPanel(
        id = "select",
        tabPanel("Chemostat", {
           renderPlot({
             # set parameters
             params <- c(mumax = input$mumax, Vmax = input$Vmax, Km = input$Km, Qmin = input$Qmin, Rs = input$Rs, d = input$d)
             state <- c(R = input$R, Q = input$Q, X = input$X)
        
            # generate model output
            times <- seq(0, input$t_final, length = 100)
            out <- ode(y = state, times = times, func = culture.chemostat, parms = params) %>% as.data.frame()
        
            # plot time series
            plotTS(out)
          }, width=w, height=h)
        }),
        tabPanel("Turbidostat", {
           renderPlot({
             # set parameters
             params <- c(mumax = input$mumax, Vmax = input$Vmax, Km = input$Km, Qmin = input$Qmin, Rs = input$Rs, Xstar = 10^(input$log10Xstar))
             state <- c(R = input$R, Q = input$Q, X = input$X)
        
            # generate model output
            times <- seq(0, input$t_final, length = 100)
            out <- ode(y = state, times = times, func = culture.turbidostat, parms = params) %>% as.data.frame()
        
            # plot time series
            plotTS(out)
          }, width=w, height=h)
        }),
        tabPanel("Batch", {
           renderPlot({
             # set parameters
             params <- c(mumax = input$mumax, Vmax = input$Vmax, Km = input$Km, Qmin = input$Qmin, Rs = input$Rs, d = 0)
             state <- c(R = input$R, Q = input$Q, X = input$X)
        
            # generate model output
            times <- seq(0, input$t_final, length = 100)
            out <- ode(y = state, times = times, func = culture.chemostat, parms = params) %>% as.data.frame()
        
            # plot time series
            plotTS(out)
          }, width=w, height=h)
        }),
        tabPanel("Semi-continuous", {
           # run a series of batch cultures of length Period and diluting by a factor Df at the end of each
           renderPlot({
             # set parameters
             params <- c(mumax = input$mumax, Vmax = input$Vmax, Km = input$Km, Qmin = input$Qmin, Rs = input$Rs, d = 0)
             state <- c(R = input$R, Q = input$Q, X = input$X)
        
             # generate model output
             t0 = 0
             times <- seq(t0, t0+input$Period, length = 100)
             out <- ode(y = state, times = times, func = culture.chemostat, parms = params) %>% as.data.frame()
             while(out$time[nrow(out)] < input$t_final) {
               t0 <- out$time[nrow(out)]
               times <- seq(t0, t0+input$Period, length = 100)
               Df = input$Xstar/out$X[nrow(out)]
               state <- c(R = Df * out$R[nrow(out)] + (1-Df)*input$Rs,
                          Q = out$Q[nrow(out)],
                          X = Df * out$X[nrow(out)] + 0) 
               out <- bind_rows( out, ode(y = state, times = times, func = culture.chemostat, parms = params) %>% as.data.frame() )
             }
        
             # plot time series
             plotTS(out)
          }, width=w, height=h)
          })
      )
    )
  )
)
        
     

```


## Non-dimensionalized equations

Our model has one independent variable ($t$, time), three dependent (or "state") variables ($R$, $Q$, and $X$) and 6 parameters ($\mu_{max}$, $d$, $Q_{min}$, $V_{max}$, $K_m$, and $R_s$.) Some of these parameters are best understood in terms of their values realtive to each other. If the dilution rate is larger than the maximum growth rate, then washout is inevitable, so a natural scale for $\mu_{max}$ is its magnitude relative to $d$. Both of these parameters have the same dimensions (Time$^{-1}$) so their quotient $\tilde\mu_{max}=\mu_{max}/d$ is dimensionless. In our model we are interested in any value $0 < \tilde\mu_{max} < 1$. Similarly, the half saturation constant and resource supply concentration both have the same dimensions (mol Volume$^{-1}$, although mol is not really a dimension, but it does represent the measurement scale of the amount of resource). We can form a dimensionless value $\tilde R_s = R_s/K_m$ and fix $K_m=1$. We are interested in any positive value of $\tilde R_s$, but values less than 1 will lead to strongly resource limiting situations. The third pair, minimum cell quota and maximum uptake rate have different dimensions (mol cell$^{-1}$ and mol cell$^{-1}$ Time$^{-1}$). The first we can fix at one and use it to redefine cell quota, $\tilde Q = Q / Q_{min}$. The second we scale by both $d$ and $Q_{min}$ to get the following dimensionless quantity: $\tilde V_{max} = V_{max}/(Q_{min}d)$. After all this, we have fixed three parameters at the value 1 and so we really now only have three parameters in our equations. We need to scale all our other variables to make them dimensionless, so we write $\tilde R = R/K_m$, $\tilde X = X Q_{min}/K_m$, and $\tilde t = td$. We substitute these 7 equations into our original model to change the equations into dimensionless quantities. In keeping with our notation all variables below should have a tilde decoration, but for simplicity we omit it:

$$\frac{dR}{dt} = (R_s - R) - X\left(\frac{V_{max}{R}}{1+R} \right)$$

$$\frac{dQ}{dt} = \frac{V_{max}R}{1+R} - \mu_{max}(Q-1)$$
$$\frac{dX}{dt} = X\left(\mu_{max}\left(1-\frac{1}{Q}\right) - 1\right)$$
These are exactly the same as our original equations but with dimensionless quantities measured relative to $d=1$, $Q_{min}=1$ and $K_m=1$. Incidentally, all this analysis explains why it is not important to use biologically meaningful values in our analyses presented above.

Since total resource mass in the culture vessel is $M = R + XQ$, dimensionless mass is $\tilde M = \frac{M}{K_m} = \tilde R + \tilde X \tilde Q$.

Since $d=1$, this transformation does not work for batch cultures or semi-continuous batch cultures. Instead fix $\mu_{max}=1$.


## Biologically plausible values

A parameter-fitting exercise, using the data in [Liefer et al (2019)](https://dx.doi.org/10.3389/fmicb.2019.00763) yields the following estimates for growth rate and minimum N cell quota for _Thalassiosira pseudonana_.

$\mu_{max} = 0.70 d^{-1}$, $Q_{min} = 0.037$ pmol cell$^{-1}$.

Data from that study do not constrain $V_{max}$ or $K_m$ well, but $K_m$ appears to be small and $V_{max}$ approximately matches growth rate. For kinetic studies of uptake of nitrate, nitrite, phosphate, and silicate see [Parslow et al (1984)](https://link.springer.com/content/pdf/10.1007/BF00393085.pdf).

We will use $V_{max} = 1.0 d^{-1}$ and $k_m = 0.1 \mu$ mol L$^{-1}$. Note that $V_{max}$ is reported relative to $Q$ in Parslow et al (1984).

This simple model does not place an upper bound on $Q$.


## Estimating growth rates

A good model should provide clarity about the meaning of measurements and the ability to make predictions. Since growth rate is a key quantity in these models, it's natural to ask what we can learn about estimating growth rates.

In a batch cutlure, the conditions are always changing, so the growth rate is changing. The model can be fit to observed data and used to estimate growth rate as it changes over time. This is not commonly done and the interpretation of the result depends on the model being at least partly correct.

In a chemostat, the growth rate is equal to the dilution rate set by the experimentalist, once equilibrium has been achieved.

In a turbidostat, the growth rate is the dilution rate, but the experimental control is the cell density. The dilution rate (volume of new media added per day divided by the constant culture volume) is measured as the dilution is tracked by the monitoring software. In our turbidostat culture is added every minute as needed, which is a very good approximation to a continuous culture. So the calculation of growth rate is easy: dilution rate per unit time. Time periods of any length can be used with the caveat that the cell density should be constant over the observed time period. If conditions, e.g., irradiance, are changing, then growth rate will be changing and we can only measure average growth rate. Fancier analysis is possible to estimate instantaneous growth rate using data assimilation methds.

Finally, the semi-continuous batch growth rate, which is most familiar to people in our lab, is best estimated from linear regression of log projected cell density (cell density scaled by periodic, discrete dilutions) against time. How and why is this different from the turbidostat growth rate? Why can't we just use dilution rate or why don't we use linear regression for the turbidostat? The first reason is the continuous culture has discrete dilutions which are infrequent compared to the growth rate (there may be multiple doublings between dilutions) and the second reason is that the biomass is not constant in a continuous culture. Linear regression can be used in a turbidostat to estimate growth rate, by using the frequent dilutions to scale the cell densities. It simply isn't necessary.

## Exercises (not developed yet)

Play around with the various scenarios.

* Semi-continuous batch: Find a set-up where $\mu$ is approximately constant after a few dilutions. What happens if you make the time between dilutions longer? What seems to change faster, the growth rate or the slope on the $\log X(t)$ vs $t$ curve?

* Check the dimensionless quantity reasoning. Set $d=1$, $Q_{min}=1$, and $K_m=1$ and note the output of the model. Then reduce each by a factor of two and scale the other three constants and the inital values. You should obtain the same results, except for a scaling factor of 2 in time and total mass.

* Check that mass is conserved (to within a few parts per million; the error arises from the numerical method to solve the equations) once the system comes to a steady state regardless constants (the values of $R_s$ and initial conditions do matter). If the initial mass in the reaction vessel is very close to the supply concentration, then the very small and noisy deviations will be visible in the trajectory of $M$.
